Inferring neural networks from fluorescent calcium imaging using partial correlation and sample weighting

### 2014-05-30 article.tex
Introduce gaussian hypothesis

% introduce the notion of dependence / independence
% TODO add a more formal definition of independence.
From a probabilistic point of view, the inference of the undirected network
can be formulated as inferring conditional dependences and
independences. Each neuron signal $i$ is a random continuous variable $X_i$,
which follows a probability density function $p_\mathcal{X_i}(x_i)$.
Two neurons $X_i$, $X_j$ will be said conditionally independent
$X_i \perp X_j | X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}$ to all
others neuron if
\begin{align*}
&p_{X_i, X_j | X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}}
    (x_i, x_j | x_1,\ldots,x_p \backslash \left\{x_i, x_j\right\}) \\
=&p_{X_i | X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}}
    (x_i | x_1,\ldots,x_p \backslash \left\{x_i, x_j\right\})
p_{X_j | X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}}
    (x_j | x_1,\ldots,x_p \backslash \left\{x_i, x_j\right\})
\end{align*}
for all real numbers $x_1,\ldots,x_p $ such that
$p_{X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}}
 (x_1,\ldots,x_p \backslash \left\{x_i, x_j\right\}) > 0$.
Thus if $X_i \not\perp X_j | X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}$,
then we have the edge $i-j$ in the undirected graph.

% Introduce correlation => Estimator of I(X; Y)
%                       => Issue indirect effect
% Introduce partial correlation => Estimator of I(X; Y | Z)
%                               => Symmetric measure
% Introduce a custom solution to make the Y matrix assymetric
% Introduce Filtering

@@

A common association measure is the Pearson correlation coefficient.
\begin{align}
\rho_{ij} = \dfrac{cov(I,J)}{\sigma_I \sigma_J}
\end{align}
Even if \textit{correlation does not imply causation} [todo cite], it may give
an insight on the underlying connectivity network. Nevertheless, the main
challenge in inferring a network is to distinguish direct from indirect
interactions [de la Fuente et al., 2004]. While the classical correlation
coefficient characterize both direct and indirect interactions, the partial
correlation coefficient quantifies association between variables conditionally
on one or several other variables (the number of variables considered being
the order of the partial correlation coefficient). [Shipley, 2002] suggests to
consider correlation coefficients of all possible orders, i.e. conditioned
on every subsets of the set of $p-2$ other variables, in order to take into
account multiple indirect paths between a given pair of variables.  An obvious
limitation to calculate high-order coefficient is the number of samples in
conditioned data sets. Considering the large number of samples at our disposal
in the connectomics challenge, we only consider the highest order of partial
correlation coefficient. We thus try to estimate partial correlation matrix as
the inferred network where a matrix element, denoted by
$\rho_{ij | \mathcal{V}^{-i,j}}$, is the partial correlation coefficient of
order $p-2$ between neurons $i$ and $j$. Note that this matrix is symmetric
and thus it is impossible to determine the causal mechanism behind the
interaction (I directly causes J, J directly causes I or both
[de la Fuente, 2004])\footnote{Two other causal mechanisms may be implied
by a non-zero partial coefficient. First, the pair of variables may be induced
by a common (hidden) variable and second, I (respectively J) is conditionally
correlated to a hidden variable affecting J (resp. I) [de la Fuente, 2004].} .
A way to find these coefficient [todo cite?] is to take the inverse of the
covariance matrix to get the partial correlation matrix of order $p-2$, also
known as the precision matrix [todo cite] .

>>>>

% introduce the notion of dependence / independence
From a probabilistic point of view, the inference of the undirected network
can be formulated as inferring conditional dependences and
independences. Each neuron signal $i$ is a random continuous variable $X_i$,
which follows a probability density function $p_\mathcal{X_i}(x_i)$.
Two neurons $X_i$, $X_j$ will be said conditionally independent
to a set of neurons $X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}$, denoted
by  $X_i \perp X_j | X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}$,  if
\begin{align*}
&p_{X_i, X_j | X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}}
    (x_i, x_j | x_1,\ldots,x_p \backslash \left\{x_i, x_j\right\}) \\
=&p_{X_i | X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}}
    (x_i | x_1,\ldots,x_p \backslash \left\{x_i, x_j\right\})
p_{X_j | X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}}
    (x_j | x_1,\ldots,x_p \backslash \left\{x_i, x_j\right\})
\end{align*}
for all real numbers $x_1,\ldots,x_p $ such that
$p_{X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}}
 (x_1,\ldots,x_p \backslash \left\{x_i, x_j\right\}) > 0$.
Thus if $X_i \not\perp X_j | X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}$,
then we have the edge $i-j$ in the undirected graph.

Under the hypothesis that the neurons follow a jointly Gaussian distribution
$p_{X_1,\ldots,X_p}(x_1,\ldots,x_p) \sim \mathcal{N}(\mu; \Sigma)$
of mean vector $\mu$ and of covariance matrix $\Sigma$, two neurons $X_i$
and $X_j$ are independent $X_i \perp X_j$ if they have zero correlation
coefficient and are conditionally independent to all others neurons
$X_i \perp X_j | X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}$ if they
have zero partial correlation coefficient. Even if \textit{independence
does no imply causation}, it gives insight on the underlying connectivity
network.

% Introduce correlation => Estimator of I(X; Y) => Issue indirect effect
The Pearson correlation coefficient $\rho_{ij}$ is an association measure
and defined by
\[
\rho_{ij} = \frac{\Sigma_{ij}}{\sqrt{\Sigma_i \Sigma_j}}.
\]

% Introduce partial correlation => Estimator of I(X; Y | Z)
%                               => Symmetric measure

% Introduce a custom solution to make the Y matrix asymmetric
% Introduce Filtering

@@

A common association measure is the
\begin{align}
\rho_{ij} = \dfrac{cov(I,J)}{\sigma_I \sigma_J}
\end{align}
Even if \textit{correlation does not imply causation} [todo cite], it may give
an insight on the underlying connectivity network. Nevertheless, the main
challenge in inferring a network is to distinguish direct from indirect
interactions [de la Fuente et al., 2004]. While the classical correlation
coefficient characterize both direct and indirect interactions, the partial
correlation coefficient quantifies association between variables conditionally
on one or several other variables (the number of variables considered being
the order of the partial correlation coefficient). [Shipley, 2002] suggests to
consider correlation coefficients of all possible orders, i.e. conditioned
on every subsets of the set of $p-2$ other variables, in order to take into
account multiple indirect paths between a given pair of variables.  An obvious
limitation to calculate high-order coefficient is the number of samples in
conditioned data sets. Considering the large number of samples at our disposal
in the connectomics challenge, we only consider the highest order of partial
correlation coefficient. We thus try to estimate partial correlation matrix as
the inferred network where a matrix element, denoted by
$\rho_{ij | \mathcal{V}^{-i,j}}$, is the partial correlation coefficient of
order $p-2$ between neurons $i$ and $j$. Note that this matrix is symmetric
and thus it is impossible to determine the causal mechanism behind the
interaction (I directly causes J, J directly causes I or both
[de la Fuente, 2004])\footnote{Two other causal mechanisms may be implied
by a non-zero partial coefficient. First, the pair of variables may be induced
by a common (hidden) variable and second, I (respectively J) is conditionally
correlated to a hidden variable affecting J (resp. I) [de la Fuente, 2004].} .
A way to find these coefficient [todo cite?] is to take the inverse of the
covariance matrix to get the partial correlation matrix of order $p-2$, also
known as the precision matrix [todo cite] .
